---
title: "data-preprocessing"
author: "Bowen He"
date: "7/26/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(raster)
library(sf)
library(exactextractr)
library(sp)
library(tidycensus)
library(tidyverse)
library(Metrics)
library(stars)
library(rprojroot)
library(rgeos)
proj_root <- find_root(is_rstudio_project)
CRS <- "+proj=utm +zone=16 +datum=NAD83 +units=km +no_defs +ellps=GRS80 +towgs84=0,0,0"
```

In this Rmarkdown document, I will read in the decennial 2020 census tract dataframe, 
decennial 2020 census block group dataframe, decennial 2020 census block dataframe, 
and the 2019 NLCD land cover raster.

Then, I will perform the spatial data join analysis to integrate the decennial 2020 CT, 
the decennial 2020 CBG, the decennial 2020 CB, and the 2019 NLCD land cover data together.
The product (ct_land_cover_df.Rds) of this R code is a dataframe that contains each 
2020 decennial census tract of the Davidson County, Nashville and its corresponding population, 
sum of each land cover codes as well as the census tract it belongs to.

This dataframe will be used to perform the regularized analysis to select the significant land
cover covariates in the proposed spatial model.

# Retrieve Decennial 2020 total population data {#decennial2020-pop}
Retrieve the Decennial 2020 total population data of the Davidson County, Nashville
as the research subject of interest.
```{r ct-process include = FALSE, message = FALSE, warning = FALSE}
pop.tract <- get_decennial(geography = "tract", sumfile = "pl", state = 'TN', county = 'Davidson', variables = "P1_001N", year = 2020, geometry = TRUE)
census_tract <- pop.tract %>%
  dplyr::select(-variable, -value) %>% rename('CT_NAME' = 'NAME') %>%
  mutate(area = st_area(.))

st_write(census_tract, dsn = file.path(proj_root, "data", 'src/census_tract'), layer = "census_tract",
         driver = "ESRI Shapefile", append = FALSE)
# CBG
pop.cbg <- get_decennial(geography = "block group", sumfile = 'pl', state = 'TN', county = 'Davidson', variables = "P1_001N", year = 2020, geometry = TRUE)
census_block_group <- pop.cbg %>%
  dplyr::select(-variable, -value) %>% rename('CBG_NAME' = 'NAME')%>%
  mutate(area = st_area(.))

st_write(census_block_group, dsn = file.path(proj_root, "data", 'src/census_block_group'), layer = "census_block_group",
         driver = "ESRI Shapefile", append = FALSE)

# CB
pop.cb <- get_decennial(geography = "block", sumfile = 'pl', state = 'TN', county = 'Davidson', variables = "P1_001N", year = 2020, geometry = TRUE)
census_block <- pop.cb %>% 
  dplyr::select(-variable, -value) %>% rename('CB_NAME' = 'NAME')%>%
  mutate(area = st_area(.))

st_write(census_block, dsn = file.path(proj_root, "data", 'src/census_block'), layer = "census_block",
         driver = "ESRI Shapefile", append = FALSE)
```

# spatially join the CB and CT dataframe {#CB-join-CT}
Create another column called TRACT to include the information of the census tract that the census block belongs to.
```{r spatial-join-cb-ct include = FALSE, message=FALSE, warning=FALSE}
CB_CT <- CB %>% 
  st_drop_geometry() %>%
  extract(CB_NAME, into = c('BLOCK', 'TRACT'), "^(Block \\d+).*(Census Tract.*)") %>%
  left_join(CB %>% dplyr::select(-CB_NAME, -area), by = 'GEOID') # re

st_write(CB_CT, dsn = file.path(proj_root, "data", 'src/cb_ct'), layer = "cb_ct",
         driver = "ESRI Shapefile", append = FALSE)
```

# Import the 2019 NLCD raster {#nlcd-raster}
Import the newest 2019 Davidson county NLCD land cover raster data.
```{r 2019-nlcd include = FALSE, message = FALSE, warning = FALSE}
nlcd_raster <- raster::raster(file.path(proj_root, "data/src/NLCD-2019/nlcd-2019.img"))
```

# Transform 2019 NLCD raster points{#nlcd-pts}
```{r 2019-nlcd-pts include = FALSE, message = FALSE, warning = FALSE}
nlcd_pts <- rasterToPoints(nlcd_raster) %>% 
  as.data.frame() %>%
  st_as_sf(coords = c('x', 'y')) %>%
  st_set_crs(CRS)

write_rds(list(nlcd_pts = nlcd_pts),
          path = file.path(data_dir = file.path(proj_root, "data"), "src/nlcd_pts.Rds"))
```

# Spatial join the NLCD raster with the census tract {#ct-sum-nlcd}
Sum the land cover grids of each land cover type into Davidson county shapefile: census tract.
```{r spatial-sum-nlcd include = FALSE, message=FALSE, warning=FALSE}
#spatial join the nlcd raster (2019) with the census tract
ct_land_cover_df <- st_join(nlcd_pts, census_tract %>% st_transform(crs = CRS), join = st_intersects) %>%
  st_drop_geometry() %>%
  group_by(GEOID) %>%
  count(nlcd.2019) %>%
  pivot_wider(names_from = nlcd.2019, values_from = n)

ct_land_cover_df <- filter(ct_land_cover_df, !is.na(GEOID))
ct_land_cover_df[is.na(ct_land_cover_df)] <- 0 
ct_land_cover_df <- ct_land_cover_df %>%
   dplyr::select(-`128`)

land_use_table <- c(
  "GEOID" = "GEOID",
  "21" = "developed_open_space",
  "22" = "developed_low_intensity",
  "23" = "developed_medium_intensity",
  "24" = "developed_high_intensity",
  "31" = "barren",
  "41" = "deciduous_forest",
  "42" = "evergreen_forest",
  "43" = "mixed_forest",
  "52" = "shrub",
  "71" = 'grassland',
  "81" = 'pasture',
  "11" = "open_water",
  "82" = "cultivated_crops",
  "90" = "woody_wetlands",
  "95" = "herbaceous_wetlands"
)

colnames(ct_land_cover_df) <- land_use_table[colnames(ct_land_cover_df)] 

ct_land_cover_df <- left_join(census_tract, ct_land_cover_df, by = 'GEOID') %>%
  mutate(area = st_area(.)) %>%
  drop_units()

write_rds(list(ct_land_cover_df = ct_land_cover_df),
          path = file.path(data_dir = file.path(proj_root, "data"), "src/ct_land_cover_df.Rds"))
```

# spatial join the census tract and the NLCD grids {ct-join-nlcd} 
```{r spatial-join-nlcd include = FALSE, message=FALSE, warning=FALSE}
#spatial join the nlcd raster (2019) with the census tract
nlcd_grids_ct_df <- st_join(nlcd_pts, census_tract %>% st_transform(crs = CRS), join = st_intersects) %>%
   st_drop_geometry() 

# append the land use information
nlcd_grids_ct_df <- nlcd_grids_ct_df %>%
   mutate(land_use = land_use_table[as.character(nlcd_grids_ct_df$nlcd.2019)])
 
# generate the joined census tract and NLCD grids df 
write_rds(list(nlcd_grids_ct_df = nlcd_grids_ct_df),
           path = file.path(data_dir = file.path(proj_root, "data"), "src/nlcd_grids_ct_df.Rds"))
```
